---
title: "Lab 0: Getting Started with dplyr"
subtitle: "Your First Data Analysis"
author: "Jenny Brar"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Overview

Welcome to your first lab! In this (not graded) assignment, you'll practice the fundamental dplyr operations I overviewed in class using car sales data. This lab will help you get comfortable with:

- Basic data exploration
- Column selection and manipulation  
- Creating new variables
- Filtering data
- Grouping and summarizing

**Instructions:** Copy this template into your portfolio repository under a `lab_0/` folder, then complete each section with your code and answers. You will write the code under the comment section in each chunk.
Be sure to also copy the data folder into your `lab_0` folder.

# Setup

```{r setup}
# Load the tidyverse library
library(tidyverse)

# Read in the car sales data
# Make sure the data file is in your lab_0/data/ folder
car_data <- read_csv("data/car_sales_data.csv")
```

# Exercise 1: Getting to Know Your Data

## 1.1 Data Structure Exploration

Explore the structure of your data and answer these questions:

```{r explore-structure}
# Use glimpse() to see the data structure
glimpse(car_data)


# Check the column names


# Look at the first few rows
head(car_data)
```

**Questions to answer:**
- How many rows and columns does the data set have? 7 columns and 50k rows. But the tibble only shows the first 6 rows. 
- What types of variables do you see (numeric, character, etc.)? There are categorical and numerical data in each row. 
- Are there any column names that might cause problems? Why? Year of Manufacturer is quite long 

**Your answers:**
- Rows: 50k
- Columns: 7  
- Variable types: Numerical and Categorical
- Problematic names: Year of Manufacturer

## 1.2 Tibble vs Data Frame

Compare how tibbles and data frames display:

```{r tibble-vs-dataframe}
# Look at the tibble version (what we have)
print(car_data)

# Convert to regular data frame and display
car_df <- as.data.frame(car_data)
print(car_df)
```

**Question:** What differences do you notice in how they print?

**Your answer:** Tibble shows the first few rows, like a preview. Data frames show the first 143 rows (it's max), and is not organized in a table format. 

# Exercise 2: Basic Column Operations

## 2.1 Selecting Columns

Practice selecting different combinations of columns:

```{r column-selection}
# Select just Model and Mileage columns
selected_column <- car_data %>%
  select("Model", "Mileage")


# Select Manufacturer, Price, and Fuel type
selected_man_pri_FT <- car_data %>%
  select("Manufacturer", "Price", "Fuel type")
selected_man_pri_FT

# Challenge: Select all columns EXCEPT Engine Size
select(car_data, -"Engine size")

```

## 2.2 Renaming Columns

Let's fix a problematic column name:

```{r rename-columns}
# Rename 'Year of manufacture' to year
car_data <- car_data %>%
  rename(year = `Year of manufacture`)


# Check that it worked
car_data

```

**Question:** Why did we need backticks around `Year of manufacture` but not around `year`?

**Your answer:** 
to help R understand that this is a column name that already exists, rather than a specific text to use. 

# Exercise 3: Creating New Columns

## 3.1 Calculate Car Age

```{r calculate-age}
# Create an 'age' column (2025 minus year of manufacture)
car_data <- car_data %>%
  mutate(age = 2025 - year)

# Create a mileage_per_year column 
car_data <- car_data %>%
  mutate(mileage_per_year = Mileage/age)

# Look at your new columns
#select(car_data, Model, year, age, Mileage, mileage_per_year)

car_data %>%
  select("Model", "year", "age", "Mileage", "mileage_per_year")
```

## 3.2 Categorize Cars

```{r categorize-cars}
# Create a price_category column where if price is < 15000, its is coded as budget, between 15000 and 30000 is midrange and greater than 30000 is mid-range (use case_when)
library(dplyr)

car_data <- car_data %>%
  mutate(
    price_category = case_when(
      Price < 15000 ~ "budget",
      Price >= 15000 & Price <= 30000 ~ "midrange",
      Price > 30000 ~ "premium"
    )
  )

# Check your categories select the new column and show it
car_data%>% 
  select(price_category)
```

# Exercise 4: Filtering Practice

## 4.1 Basic Filtering

```{r basic-filtering}
# Find all Toyota cars
car_data %>%
  filter(Manufacturer == "Toyota")


# Find cars with mileage less than 30,000
car_data %>%
  filter(Mileage < 30000)


# Find luxury cars (from price category) with low mileage
car_data %>%
  filter(
    price_category == "premium", 
    Mileage < 30000
  )

```

## 4.2 Multiple Conditions

```{r multiple-conditions}
# Find cars that are EITHER Honda OR Nissan
car_data %>%
  filter(Manufacturer== "Honda" |Manufacturer=="Nissan")

# Find cars with price between $20,000 and $35,000

car_data %>%
  filter(Price == 20000 & Price == 35000)


# Find diesel cars less than 10 years old
car_data %>%
  filter(`Fuel type`== "Diesel", 
         age < 10)

```

**Question:** How many diesel cars are less than 10 years old?

**Your answer:** 2040

# Exercise 5: Grouping and Summarizing

## 5.1 Basic Summaries

```{r basic-summaries}
# Calculate average price by manufacturer
avg_price_by_brand <- car_data %>%
  group_by(Manufacturer) %>%
  summarize(avg_price = mean(Price, na.rm = TRUE))

avg_price_by_brand

# Calculate average mileage by fuel type

avg_mileage_fuel <- car_data %>%
  group_by(`Fuel type`) %>%
  summarise(avg_mileage = mean(Mileage, na.rm = TRUE))

avg_mileage_fuel


# Count cars by manufacturer

car_count <- car_data %>%
  group_by(Manufacturer) %>%
  summarise(cars = n())

car_count

```

## 5.2 Categorical Summaries

```{r categorical-summaries}
# Frequency table for price categories
car_data %>%
  count(price_category)

car_data %>%
  count(price_category, name = "Frequency")

#Another way of doing the same thing...

table(car_data$price_category)

```


# Submission Notes

**To submit this lab:**
1. Make sure your code runs without errors
2. Fill in all the "[YOUR ANSWER]" sections and complete all of the empty code! 
3. Save this file in your portfolio's `lab_0/` folder
4. Commit and push to GitHub
5. Check that it appears on your GitHub Pages portfolio site

**Questions?** Post on the canvas discussion board or come to office hours!