---
title: "Bikeshare_SpaceTime_Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Jenny"
date: today
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# INTRODUCTION

## The Rebalancing Challenge in Philadelphia

Philadelphia's Indego bike share system faces the same operational challenge as every bike share system: **rebalancing bikes to meet anticipated demand**.

Considering there are 200 stations across Philadelphia, limited trucks and staff for moving bikes, and 2-3 hours before morning rush hour demand peaks. **The question:** this report is grounded in are if stations will run out of bikes by 8:30 AM?

This report builds predictive models that forecast bike share demand across space (different stations) and time (different hours) to help solve this  problem. The first part of the report explores data from Q1 2025 (January through March) and tests various temporal, weather, demographic, and spatial features to find the best model for predicting hourly trip demand at the station level. Through a series of progressively more complex linear models, we identify which predictors drive the most improvement in forecast accuracy. The second part applies the same workflowto Q4 2025 (October through December), which had interesting weather patterns, warm October followed by a cold snap in mid-December, as well as notable holidays including Thanksgiving, and Christmas. Three additional features were engineered, specifically holiday indicators, a nice weather feature, and distance to the nearest bike lane.

By comparing the first part of the year to the last, this report aim to understand how seasons shape ridership behaviour and model performance, if  predictors that matter most in winter carry over to fall, and what the error patterns reveal about which communities and time periods are hardest to predict. These two quarters offer a better picture of the operational and equity dimensions of deploying a bike share prediction system in Philadelphia.

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)


library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier

```

```{r themes}
#| echo: false
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set-census-api-key}
#| echo: false
census_api_key(Sys.getenv(""))

#Use the following code if you're not able to access the key in your environment
#census_api_key("yourkeyhere", install=TRUE, overwrite=TRUE)
```


## Q1 Indego Trip Data (Q1 2025)

```{r load_indego}
#| message: false
#| warning: false
# Read Q1 2025 data
indego <- read_csv(here("data/indego-trips-2025-q1.csv"))

# Quick look at the data
glimpse(indego)
```


## Examine the Data Structure

```{r explore_data}
#| message: false
#| warning: false
# How many trips?
cat("Total trips in Q1 2025:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Create Time Bins

We need to aggregate trips into hourly intervals for our panel data structure. Read up on the lubridate library for more information on the date/time data extraction.

```{r create-time-bins}
#| message: false
#| warning: false
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

# EDA OF INDEGO DATA

## Trips Over Time

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q1 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```


```{r eagles-comparison}
#| message: false
#| warning: false
fly_eagles_fly <- daily_trips %>% filter(date == "2025-02-14")

typical_boring_friday <- indego %>%
  filter(dotw == "Fri", date != "2025-02-14") %>%
  group_by(date) %>%
  summarize(trips = n()) %>%
  summarize(avg_friday_trips = mean(trips))

print(fly_eagles_fly)
print(typical_boring_friday)

```

## Hourly Patterns

```{r hourly_patterns}
#| message: false
#| warning: false
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>% #group by hour within each weekend vs weekday
  summarize(avg_trips = n() / n_distinct(date)) %>% #total trips divided by total unique dates (rn there are multiple of the same dates/ station)
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

## Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>% #how many rows fall within this combination then nameing this count as trips
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# PUT DATA IN CONTEXT

## Load Philadelphia Census Data

We'll get census tract data to add demographic context to our stations.

```{r load_census}
#| message: false
#| warning: false


# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
glimpse(philly_census)
```

## Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

## Join Census Data to Stations

We'll spatially join census characteristics to each bike station.

```{r join_census_to_stations}
#| message: false
#| warning: false
# Create sf object for stations
stations_sf <- indego %>% 
  distinct(start_station, start_lat, start_lon) %>% #keeps only the unique combos bc there is so many repeated dates and locations, this will collapse it into one row/station
  filter(!is.na(start_lat), !is.na(start_lon)) %>% #to prevent issues when we use the st_as_sf below, we have to remove all the NA values from LAT and LON columns
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry() #we have to drop the geometry here because both philly_census and stations_sf have data. one needs to be removed so it can be joined. When we do a left join here, all rows from sation_sf stay, and even if it doesnt have a match in philly_census, it will appear with an NA. 

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station" #Joins median income to each station via station name, but we have to select only Med_Inc so not everything is joined. 
  ) %>%
  mutate(has_census = !is.na(Med_Inc)) #creates a boolean column where if a station has a CT, it will state "has_census". True when there is med income, False when there isnt. 

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

## Dealing with missing data

We need to decide what to do with the non-residential bike share stations. For this example, we are going to remove them -- this is not necessarily the right way to do things always, but for the sake of simplicity, we are narrowing our scope to only stations in residential neighborhoods. We might opt to create a separate model for non-residential stations..

```{r valid-stations}
#| message: false
#| warning: false
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

## Get Weather Data

Weather significantly affects bike share demand! Let's get hourly weather for Philadelphia.

```{r get_weather}
#| message: false
#| warning: false
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q1 2025: January 1 - March 31
weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-01-01",
  date_end = "2025-03-31"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))
```

## Visualize Weather Patterns

```{r visualize-weather}
#| message: false
#| warning: false
#| 
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q1 2025",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

# CREATE SPACE-TIME PANEL

## Aggregate Trips to Station-Hour Level

```{r aggregate-trips}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

## Create Complete Panel Structure

Not every station has trips every hour. We need a **complete panel** where every station-hour combination exists (even if Trip_Count = 0).

```{r complete-panel}
#| message: false
#| warning: false
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features

```{r add_time_features}
#| message: false
#| warning: false
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```

## Join Weather Data

```{r join_weather}
#| message: false
#| warning: false
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60") #join weather by hour

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

# CREATE TEMPORAL LAG VARIABLES

The key innovation for space-time prediction: **past demand predicts future demand**.

If there were 15 bike trips from Station A at 8:00 AM, there will probably be \~15 trips at 9:00 AM. We can use this temporal persistence to improve predictions.

**Note:** We still can't use spatial lags in prediction. Temporal lags are different because time moves in one direction - past predicts the future, but the future does not predict the past. Spatial lags are non-directional and create circular prediction logic.

```{r create_lags}
#| message: false
#| warning: false
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

# TEMPORAL TRAIN/TEST SPLIT

```{r temporal-split}
#| message: false
#| warning: false
# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 10)

test <- study_panel_complete %>%
  filter(week >= 10)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

# BUILD PREDICTIVE MODELS

We'll build 5 models with increasing complexity to see what improves predictions.

## Model 1: Baseline (Time + Weather)

```{r model1}
#| message: false
#| warning: false

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

The model uses Monday as the baseline. Each coefficient represents the difference in expected trips per station-hour compared to Monday - dow_simple2 = Tuesday..

**Weekday Pattern (Tue-Fri):**

-   All weekdays have positive coefficients (0.029 to 0.052)
-   Tuesday has the highest weekday effect (+0.052)
-   Weekdays likely benefit from concentrated commuting patterns

**Weekend Pattern (Sat-Sun):**

-   Both weekend days have negative coefficients (-0.061 and -0.065)
-   This means FEWER trips per station-hour than Monday

**Hourly Interpretation**

Hour Coefficient Interpretation 0 (baseline) 0.000 trips/hour (midnight) 1 -0.018 slightly fewer than midnight ... 6 +0.151 morning activity starting 7 +0.276 morning rush building 8 +0.487 PEAK morning rush 9 +0.350 post-rush ... 17 +0.568 PEAK evening rush (5 PM!) 18 +0.389 evening declining ... 23 +0.034 late night minimal

## Model 2: Add Temporal Lags

```{r model2}
#| message: false
#| warning: false
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

## Model 3: Add Demographics

```{r model3}
#| message: false
#| warning: false
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```

## Model 4: Add Station Fixed Effects

```{r model4}
#| message: false
#| warning: false
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

**What do station fixed effects capture?** Baseline differences in demand across stations. 

## Model 5: Add Rush Hour Interaction

```{r model5}
#| message: false
#| warning: false
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

# MODEL EVALUATION

## Calculate Predictions and MAE

```{r calculate_mae}
#| message: false
#| warning: false
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison

```{r compare_models}
#| message: false
#| warning: false
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# SPACE-TIME ERROR ANALYSIS

## Observed vs. Predicted

Let's use our best model (Model 2) for error analysis.

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

## Spatial Error Patterns

Are prediction errors clustered in certain parts of Philadelphia?

```{r spatial_errors}
#| message: false
#| warning: false
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))




# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

## Temporal Error Patterns

When are we most wrong?

```{r temporal_errors}
#| message: false
#| warning: false
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```



#Q4 Analysis (Lab 5)

```{r load-indegoQ4}
#| message: false
#| warning: false

# Read Q4 2025 data
indegoQ4 <- read_csv(here("data/indego-trips-2025-q4.csv"))

# Quick look at the data
glimpse(indegoQ4)
```

```{r explore-indegoQ4}
#| message: false
#| warning: false
cat("Total Trips in Q4:", nrow(indegoQ4), "\n")

cat("Date Range:", 
    min(mdy_hm(indegoQ4$start_time)), "to", 
    max(mdy_hm(indegoQ4$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indegoQ4$start_station)), "\n")

# Trip types
table(indegoQ4$trip_route_category)

# Passholder types
table(indegoQ4$passholder_type)

# Bike types
table(indegoQ4$bike_type)

```

```{r create-time-bins-Q4}
#| message: false
#| warning: false
indegoQ4 <- indegoQ4 %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indegoQ4 %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

## EDA for Q4

### Trips Over Time

```{r trips-over-time-Q4}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indegoQ4 %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q4 2025",
    subtitle = "Fall/Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

While ridership overall seems to be declining as we head in to the colder winter months, we can see a dramatic drop in mid-October, and Mid-Decemeber. From my recollection, Mid-december was one of our first snowfalls for the winter. Some holidays like thanksgiving may justify the peak at the end of November.

### Hourly Patterns

```{r hourly-patterns-Q4}
#| message: false
#| warning: false
# Average trips by hour and day type
hourly_patterns <- indegoQ4 %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```

Interestingly, there are more peaks in ridership during the week, especialy what seems to be the start of the work-day around 8 am and at the end of the day around 17:00/5pm. While weekend ridership follows this general trend, it does not have any strong peaks. Ridership during the weekend is highest between 10 am and 4-5 pm.

### Get Weather Data

```{r get-weather-Q4}
#| message: false
#| warning: false

weather_Q4 <- riem_measures(
  station = "PHL", 
  date_start = "2025-10-01",
  date_end = "2025-12-31")

weather_Q4processed <- weather_Q4 %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),#floor date rounds the timestamp to nearest unit of time, which for us is the hour behind. 
    Temperature = tmpf,  # Temperature in F
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots. 1 nt = about 1.15 miles/hour
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

#missing any hours? If so, we'll have to use the interpolate function
weather_completeQ4 <- weather_Q4processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>% #creates seq between the earliest and the latest timestamp, while complete puts in an empty row for any missing hour. it will have NA for temp, percip, and wind speed. 
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down") #this is the actual interpolation, it will use the same value as before. Assumption that weather conditions won't change too quickly

# Look at the weather
summary(weather_completeQ4 %>% select(Temperature, Precipitation, Wind_Speed))

```

### Join Census Data to Stations

```{r join-census-Q4}
#| message: false
#| warning: false
#using the same Philly census from above
philly_censusQ4 <- philly_census


# Create sf object for stations
stations_sfQ4 <- indegoQ4 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_censusQ4 <- st_join(stations_sfQ4, philly_censusQ4, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_mapQ4 <- indegoQ4 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_censusQ4 %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data

valid_stationsQ4 <- stations_censusQ4 %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

indego_censusQ4 <- indegoQ4 %>%
  filter(start_station %in% valid_stationsQ4) %>%
  left_join(
    stations_censusQ4 %>%
      select(start_station, Med_Inc, Percent_Taking_Transit,
             Percent_White, Total_Pop),
    by = "start_station"
  )

# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme

```

## Create Space-Time Panel

```{r aggregate-trips-Q4}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panelQ4 <- indego_censusQ4 %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panelQ4)

# How many unique stations?
length(unique(trips_panelQ4$start_station))

# How many unique hours?
length(unique(trips_panelQ4$interval60))

hist(trips_panelQ4$Trip_Count)

```

```{r complete-panel-Q4}
#| message: false
#| warning: false
# Calculate expected panel size
n_stationsQ4 <- length(unique(trips_panelQ4$start_station))
n_hoursQ4 <- length(unique(trips_panelQ4$interval60))
expected_rowsQ4 <- n_stationsQ4 * n_hoursQ4

cat("Expected panel rows:", format(expected_rowsQ4, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panelQ4), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rowsQ4 - nrow(trips_panelQ4), big.mark = ","), "\n")

# Create complete panel
study_panelQ4 <- expand_grid(
  interval60 = unique(trips_panelQ4$interval60),
  start_station = unique(trips_panelQ4$start_station)
) %>%
  # Join trip counts
  left_join(trips_panelQ4 %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panelQ4)
hist(study_panelQ4$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributesQ4 <- trips_panelQ4 %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributesQ4)

study_panelQ4 <- study_panelQ4 %>%
  left_join(station_attributesQ4, by = "start_station")

glimpse(study_panelQ4)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panelQ4), big.mark = ","), "\n")



```

Compared to Q1, as expected in colder months, there are substantially less trips counts.

```{r add-time-features-Q4}
#| message: false
#| warning: false
study_panelQ4 <- study_panelQ4 %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panelQ4)
```

```{r join-weather-Q4}
#| message: false
#| warning: false
study_panelQ4 <- study_panelQ4 %>%
  left_join(weather_completeQ4, by = "interval60")
```

```{r create-lags-Q4}
#| message: false
#| warning: false
# Sort by station and time first 
study_panelQ4 <- study_panelQ4 %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panelQ4 <- study_panelQ4 %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows where lag1day is NA
# (first 24 hours of each station have no lag data yet)
study_panel_completeQ4 <- study_panelQ4 %>%
  filter(!is.na(lag1day))

```

```{r check-weeks-Q4}
#| echo: false
#have to check the weeks first:
unique(study_panel_completeQ4$week)
```

## Temporal Train/Test Split

```{r temporal-split-Q4}
#| message: false
#| warning: false
#Same thing as earlier, but now creating the training and testing dataframes for later use on q4 weeks
early_stations <- study_panel_completeQ4 %>%
  filter(week < 49) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_completeQ4 %>%
  filter(week >= 49) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

common_stations <- intersect(early_stations, late_stations)

study_panel_completeQ4 <- study_panel_completeQ4 %>%
  filter(start_station %in% common_stations)

train <- study_panel_completeQ4 %>%
  filter(week < 49)

test <- study_panel_completeQ4 %>%
  filter(week >= 49)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")
```

## Build Predictive Models with Q4 data

### Model 1: Baseline (Time + Weather)

```{r model1-Q4}
#| message: false
#| warning: false

# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun"))) #this is making dotw a categorical variable with rankings, so that everything is compared to Monday 1, Will capture  weekly patterns

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7) #contr.treatment(7) means

# Now run the linear model. As.factor ensures that the hours 0-23 remain numbers with no value in-between-not continuous
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

### Model 2: Add Temporal Lags

```{r model2-Q4}
#| message: false
#| warning: false

#Temporal Lags as predictors- LM cant account for this, which is why we use temporal lag
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```

### Model 3: Add Demographics

```{r model3-Q4}
#| message: false
#| warning: false

#adding demographics

model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)


```

### Model 4: Add Station Fixed Effects

```{r model4-Q4}
#| message: false
#| warning: false
#adding station fixed effects. Essentially, here is the dummy variable for each station. Basically, some stations are just busier than others even with the same time, weather etc. so by including as.factor(start station) all stations except the baseline 
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station), #compared to the baseline station, how many more or fewer trips does this station get. 
  data = train
)

# Summary too long with all station dummies, just show key metrics. r2 went up probably bc now we have the fixed effect to account for more variation
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

### Model 5: Add Rush Hour Interaction

```{r model5-Q4}
#| message: false
#| warning: false
#Model 5 for rush hour interaction. 
#we want to see if the effect of one variable depends on another. So, for rush hour, we want to see 

model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends- no work
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")

```

## Model Evaluation

### Calculate Predictions and MAE (Models 1-5)

```{r calculate-mae-Q4}
#| message: false
#| warning: false

#BC we want to evaluate it, we want to test it on new data

# Get predictions on test set

# Create day of week factor with treatment (dummy) coding. recreate dotw simple on test data set- it was only on train before this., 
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

#generating predictions
test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE), 
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


```

```{r compare-models-Q4}
#| message: false
#| warning: false
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


The smallest MAE is the model with temporal lags, indicating that there are fewer prediction errors when considering the data from some given hour, or time earlier.

## Error Analysis

###Spatial patterns

    ```{r error-setup-Q4}
    #| message: false
    #| warning: false

    test <- test %>%
      mutate(
        error = Trip_Count - pred2,
        abs_error = abs(error),
        time_of_day = case_when(
          hour < 7 ~ "Overnight",
          hour >= 7 & hour < 10 ~ "AM Rush",
          hour >= 10 & hour < 15 ~ "Mid-Day",
          hour >= 15 & hour <= 18 ~ "PM Rush",
          hour > 18 ~ "Evening"
        )
      )

    ```

```{r spatial-errors-Q4}
#| message: false
#| warning: false
# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors  
p1 <- ggplot() + geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5, alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma", name = "MAE (trips)", direction = -1,
    breaks = c(0.5, 1.0, 1.5), labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme

# Map 2: Average Demand
p2 <- ggplot() + geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5, alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis", name = "Avg Demand (trips/hour)", direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme

(p1 | p2)
 

```

###Temporal patterns

```{r temporal-errors-Q4}
#| message: false
#| warning: false
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


###Demographic patterns

```{r errors-demographics-Q4}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc)) #dont include stations that dont have median income

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())

```

The spatial error analysis shows that stations in Center City and along the Market-Frankford corridor carry the most prediction errors in Q4, the same pattern in Q1. High-demand stations with variable ridership from density and foot traffic is the hardest to predict regardless of season. For Temporal, weekday AM and PM rush hours generate the largest errors, consistent with Q1. However, Q4 has slightly more errors during weekend mid-day hours in October compared to Q1, possibly because mild fall weekend weather drew recreational riders in patterns not well captured by the model. The demographic error analysis replicates the Q1 finding: errors increase with neighborhood income and percent white, and decrease with transit usage. The model is more accurate in transit-dependent, lower-income communities where commuting patterns could be predictable, and less so in wealthier neighborhoods. This pattern holding across both quarters ensures confidence that it reflects a structural feature of the data instead of it being merely noise.


##Additional features to consider
###Holidays and Warma and Cold Weather

```{r feature-engineering-Q4}
#| message: false
#| warning: false
#considering the end of the year has a lot of holidays, i think a key feature to add would be a few of these:

#Weather is also key during the end of the year, especially since we had very earm weather until end of November. 

study_panel_completeQ4.1 <- study_panel_completeQ4 %>%
  mutate(
    holiday = ifelse(date %in% as.Date(c(
      "2025-10-14",
      "2025-11-27",
      "2025-11-28",
      "2025-12-25",
      "2025-12-31"
    )), 1, 0),
    too_cold = ifelse(Temperature < 32, 1, 0),
    nice_weather = ifelse(Temperature >= 55 & Precipitation == 0, 1, 0)
  )

# Re-split train and test get  new columns
train <- study_panel_completeQ4.1 %>% filter(week < 49)
test <- study_panel_completeQ4.1 %>% filter(week >= 49)


```


### Model 6: Holiday and Nice Weather

```{r model6-Q4}
#| message: false
#| warning: false
#building on model 2, which had the lowest MAE
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
contrasts(train$dotw_simple) <- contr.treatment(7)

model6 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    holiday + too_cold + nice_weather,
  data = train
)

summary(model6)


```

###Adding bike lanes

```{r load-bike-lanes}
#| message: false
#| warning: false
#there was no major improvement, so we're going to use bikelanes instead
library(sf)

# Load Philadelphia bike lanes from OpenDataPhilly

bike_lanes <- st_read("https://opendata.arcgis.com/datasets/b5f660b9f0f44ced915995b6d49f6385_0.geojson")

```

###Spatial Features
```{r calc-bike-lane-dist}
#| message: false
#| warning: false
stations_sf_project <- stations_sfQ4 %>%
  st_transform(crs = 2272)

bike_lanes_proj <- bike_lanes %>%
  st_transform(crs=2272)
#distance between each station to closet bike lane, but we want the lesser number

dist_bike <- st_distance(stations_sf_project, bike_lanes_proj)


#Ai assistance- we need to get the smallest distances in ft
stations_sfQ4$dist_bike_lane <- apply(dist_bike, 1, min)

#before we can join to panel again, we have to drop the geometry since there already is one
bike_lane_dist <- stations_sfQ4 %>%
  st_drop_geometry() %>%
  select(start_station, dist_bike_lane)
```

```{r join-bike-lane-dist}
#| message: false
#| warning: false
#now we can join it back to the study panel

study_panel_completeQ4.1 <- study_panel_completeQ4.1 %>%
  left_join(bike_lane_dist, by="start_station")
```

### Model 7: Add Bike Lane Distance

```{r model7-Q4}
#| message: false
#| warning: false
#now apply to model 2
train <- study_panel_completeQ4.1 %>% filter(week < 49)
test <- study_panel_completeQ4.1 %>% filter(week >= 49)

train <- train %>%
  mutate(dotw_simple = factor(dotw, 
    levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
contrasts(train$dotw_simple) <- contr.treatment(7)

model7 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    holiday + nice_weather + dist_bike_lane,
  data = train
)

summary(model7)

```

Better, but not significantly better

### Model 8: Poisson Regression

```{r model-poisson-Q4}
#| message: false
#| warning: false
# Recreate dotw_simple on train if needed
train <- train %>%
  mutate(dotw_simple = factor(dotw, 
    levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
contrasts(train$dotw_simple) <- contr.treatment(7)

# Poisson version of Model 2
model_poisson <- glm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train,
  family = poisson(link = "log")  #changed
)

summary(model_poisson)

```

```{r dispersion-check}
#| message: false
#| warning: false

dispersion <- sum(residuals(model_poisson, type = "pearson")^2) / 
              model_poisson$df.residual
cat("Dispersion parameter:", round(dispersion, 3), "\n")

```

The dispersion for Poisson regression is under two, which is not a severe dispersion. Variance is 1.47 times higher than predicting

```{r}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test),
    pred6 = predict(model6, newdata = test),
    pred7 = predict(model7, newdata = test),
    pred_poisson = predict(model_poisson, newdata = test, type = "response")
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction",
    "6. + Holiday & Nice Weather",
    "7. + Bike Lane Distance",
    "8. Poisson"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred6), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred7), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred_poisson), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

```{r}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison - Q4 2025",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #change angle of x-axis bc names are a bit long and  won't fit on the plot

```

```{r}

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

```

#Q4 Results and comparison
Applying the space-time prediction to Q4 2025 data produced results that were consistent with and different from Q1. For example,  Q4 dataset contained 325,361 trips across 273 residential stations between October 1 and December 31, 2025, with major holidays, and the start of winter weather.

Across all eight models, the pattern of improvement closely mirrored Q1. Model 1 (time and weather only) produced an MAE of approximately 0.51 trips per station-hour and an R² of 0.108. Adding temporal lag variables in Model 2 produced the single largest jump in performance, dropping MAE to approximately 0.41 and raising R² to 0.329. This was the best-performing model overall. Adding demographics in Model 3 offered a small additional gain in R² (0.334), while station fixed effects in Model 4 raised it further to 0.358. The rush hour interaction in Model 5 added marginal improvement at 0.361.
\
The three new features engineered for Q4 included holidays, nice weather, and bike lane distance, and  were all statistically significant. Holidays reduced expected trip counts by about 0.062 trips per station-hour (p<0.001), while nice weather days increased them by 0.048 (p<0.001), and each additional foot of distance from a bike lane reduced trips by a small but significant amount. Despite this statistical significance, neither Model 6 nor Model 7 substantially improved MAE over Model 2, suggesting that the temporal lag structure is better prediction accuracy and the new features explain variance without dramatically sharpening forecasts. The Poisson model produced a dispersion of 1.47 (mild overdispersion) and performed similarly to Model 2 for MAE. 

The most notable finding is how similar Q4 and Q1 model performance is despite the seasonal differences. Both quarters produce their best MAE of roughly 0.41 under Model 2, and ordering of model improvements is roughly the same. This suggests the underlying space-time prediction framework works well across seasons. That said, Q4 R² values are slightly higher than Q1 across all equivalent models. For example, Model 2 has R² = 0.329 in Q4 versus 0.229 in Q1. This could reflects Q4 demand being more dominated by regular commuting patterns as recreational riders drop off in colder weather, making demand easier to explain with time and weather predictors alone. Q1 by contrast includes the Eagles Super Bowl parade on February 14th, an possible outlier that increases this unexplained variance.

## Critical Reflection

###Operational implications

The best model achieves an MAE of approximately 0.41 trips per station-hour. However, it is hard to determing if this is "good enough"; it will depend on the operational context. For low-demand overnight hours, an error of 0.41 trips is negligible. For rush-hour stations in Center City that average 2–3 trips per hour, this represents roughly a 15–20% error, which could meaningfully affect rebalancing decisions. We would recommend deploying this system with human oversight rather than as a fully automated rebalancing tool. The model is most reliable for routine weekday commuting patterns and least reliable for high-demand stations during special events or unusual weather. A deployment strategy that flags high-uncertainty predictions for manual review would be best. analysis shows that prediction errors are higher in wealthier, whiter neighborhoods and lower in transit-dependent communities. This means automated rebalancing based on these predictions would be most reliable where demand is commuter-driven and least reliable in areas with more recreational or event-based riding. One concern is if errors are proportionally similar across neighborhoods, lower-income stations have lower absolute demand, so any kinbd of misallocation of bikes matters more at the margin. Safeguards should include monitoring  outcomes by neighborhood income level, ensuring that stations in lower-income areas are not systematically deprioritized when predictions are uncertain, and supplementing algorithmic decisions with local knowledge from operations people familiar with community patterns.

### limitations

The model's most significant gap is its inability to capture certain special events like concerts, sports games, protests, and festivals. These  create large unpredictable demand spikes in specific stations. It also cannot account for changes in the Indego system itself, where, for example, a  new stations opening or capacity changes. With more time and data, incorporating real-time event calendars and station capacity data could likely yeild the most meaningful improvements.




