---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Jenny Brar"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
---

# Assignment Overview

## Scenario

You are a data analyst for the ** State of Washington Department of Human Services**. The department is considering implementing an algorithmic system to identify communities that should receive priority for social service funding and outreach programs. Your supervisor has asked you to evaluate the quality and reliability of available census data to inform this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess not just what the data shows, but how reliable it is and what communities might be affected by data quality issues.

## Learning Objectives

- Apply dplyr functions to real census data for policy analysis
- Evaluate data quality using margins of error 
- Connect technical analysis to algorithmic decision-making
- Identify potential equity implications of data reliability issues
- Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an `labs/lab_1/` folder structure. Update your navigation menu to include:

```
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```
If there is a special character like a colon, you need use double quote mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidycensus)
library(tidyverse)
library(knitr)

# Set your Census API key

# Choose your state for analysis - assign it to a variable called my_state
my_state <- "Washington State"
```

**State Selection:** I have chosen Washington State for this analysis because of its interesting economic landscape. Washington State is one of the largest economies, with the Tech industry being one of the dominant ones. Yet, there is still a major pay gap, especially among women and concerns about the cost of living. Additionally, the population is majority white, with significantly smaller shares of Asian, Hispanic, and Black populations. Considering these together, it will be interesting to explore, more deeply, the underlying issues, that arise or are overlooked in technical analysis for a state that is seemingly wealthy. 

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your chosen state.

**Requirements:**
- Geography: county level
- Variables: median household income (B19013_001) and total population (B01003_001)  
- Year: 2022
- Survey: acs5
- Output format: wide

**Hint:** Remember to give your variables descriptive names using the `variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here
WA_data <- get_acs(
  geography = "county",
  variables = c(
    tot_pop = "B01003_001", 
    med_hh_inc = "B19013_001"
  ), 
  year = 2022,
  survey = "acs5",
  state = "WA", 
  output = "wide"
)
 print(WA_data)

# Makes analysis easier

# Clean the county names to remove state name and "County" 
# Hint: use mutate() with str_remove()
WA_clean <- WA_data %>%
  mutate(
    county_name = str_remove(NAME, ", Washington"),
    county_name = str_remove(county_name, "County")
  )

 
 head(WA_clean)

# Display the first few rows
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create reliability categories.

**Requirements:**
- Calculate MOE percentage: (margin of error / estimate) * 100
- Create reliability categories:
  - High Confidence: MOE < 5%
  - Moderate Confidence: MOE 5-10%  
  - Low Confidence: MOE > 10%
- Create a flag for unreliable estimates (MOE > 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
WA_reliability <- WA_clean %>%
  mutate(
    MOE_pct = round((med_hh_incM / med_hh_incE) * 100,2), 
    reliability = case_when(
      MOE_pct < 5 ~ "High Confidence",
      MOE_pct >= 5 & MOE_pct <= 10 ~ "Moderate Confidence",
      MOE_pct > 10 ~ "Low Confidence"
    )
  )

glimpse(WA_reliability)
 head(WA_data)
 
 #Identify counties with higher uncertainty based on MOE percent
 
 high_uncertainty <- WA_reliability %>%
  filter(MOE_pct > 10) %>%
  arrange(desc(MOE_pct)) %>%
  select(county_name,med_hh_incE, MOE_pct)
 
# Create a summary showing count of counties in each reliability category
 # Hint: use count() and mutate() to add percentages
 
rel_summary <- WA_reliability %>%
  group_by (reliability) %>%
  summarize(
    counties = n(), 
  )


print(rel_summary)

#Table   
 
 kable(rel_summary,
      col.names = c("Reliability", "Count of Counties"),
      caption = "Counties with Highest Income Data Uncertainty",
      format.args = list(big.mark = ","))

```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:**
- Sort by MOE percentage (highest first)
- Select the top 5 counties
- Display: county name, median income, margin of error, MOE percentage, reliability category
- Format as a professional table using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}
# Create table of top 5 counties by MOE percentage

high_uncertainty_top5 <- WA_reliability %>%
  arrange(desc(MOE_pct)) %>%
  slice(1:5) %>%
  select(county_name, med_hh_incE, med_hh_incM, MOE_pct, reliability)

# Format as table with kable()
kable(high_uncertainty_top5,
      col.names = c("County", "Median Income", "Margin of Error (MOE)", "MOE %", "Reliability"),
      caption = "Top 5 Counties with Highest Income Data Uncertainty",
      digits = 2,
      format.args = list(big.mark = ","))

# Format as table with kable() - include appropriate column names and caption
```

**Data Quality Commentary:**


Counties like Garfield, Wahkiakum, and Pend Oreille all have  Low Confidence and are most at risk of being poorly served by algorithms that may use this  data. This is concerning because, with the MOE exceeding 10%, estimates could deviate significantly from the actual values. These counties are likely small and rural, resulting in higher uncertainty. An algorithm that treats these uncertain estimates with the same confidence as high-reliability counties could improperly allocate needed funding by either overlooking communities that truly need it, or even directing resources inaccurately.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for detailed tract-level study.

**Strategy:** Choose counties that represent different reliability levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
# Store the selected counties in a variable called selected_counties
#note to self: %in% tells filter to check for values in a list

selected_counties <- WA_reliability %>%
  filter(county_name %in% c("King ", "Asotin ", "Garfield ")) %>%
  select(county_name, med_hh_incE, MOE_pct, reliability)


# Display the selected counties with their key characteristics
# Show: county name, median income, MOE percentage, reliability category
kable(selected_counties,
      col.names = c("County", "Median Income", "MOE %", "Reliability"),
      caption = "Selected Counties for Tract-Level Analysis",
      digits = 2,
      format.args = list(big.mark = ","))
```
# note to self: format.args = list(big.mark = ",")) adds commas to large numbers. better for readability. 

**Comment on the output:** Asotin County is quite close to being considered a county with higher uncertainty, but merely makes to the moderate category with an MOE of almost 9.5%. King County, where Seattle is located, has a significantly lower MOE, likely due to the large population, and also due to the prevalence of higher paying tech jobs (Seattle being the economic hub). The median income in Kings County is also significantly higher compared to the other counties. Garfield County is a small rural county with the highest uncertainty in the state. 

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected counties.

**Requirements:**
- Geography: tract level
- Variables: white alone (B03002_003), Black/African American (B03002_004), Hispanic/Latino (B03002_012), total population (B03002_001)
- Use the same state and year as before
- Output format: wide
- **Challenge:** You'll need county codes, not names. Look at the GEOID patterns in your county data for hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter

WA_demographics <- get_acs(
  geography = "tract",
  variables = c(
    total_pop = "B03002_001",
    white = "B03002_003",
    black = "B03002_004",
    hispanic = "B03002_012"
  ),
  state = "WA",
  county = c("053", "023", "033"),
  year = 2022,
  survey = "acs5",
  output = "wide"
)

head(WA_demographics)


# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations

WA_demographics <- WA_demographics %>%
  mutate(
    pct_white = (whiteE/total_popE)*100, 
    pct_black = (blackE/total_popE) *100,
    pct_hispanic = (hispanicE /total_popE) *100
  )

# Add readable tract and county name columns using str_extract() or similar
WA_demographics <- WA_demographics %>%
  mutate(
    tract_name = str_remove(NAME, "Census Tract ") %>%
                 str_remove("; Washington")
  )
```
Notes
output (wide) = one variable per row
str_remove() —  want to delete part of it
str_extract() — to pull out a specific piece from the string

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract

# Tract with highest percentage of Hispanic/Latino residents
top_hisp <- WA_demographics %>%
  arrange(desc(pct_hispanic)) %>%
  slice(1) %>%
  select(tract_name, pct_hispanic)

print(top_hisp)

# Average demographics by county..Code for extract from AI

  WA_demographics <- WA_demographics %>%
  mutate(county_name = str_extract(NAME, "[A-Za-z ]+County"))
  
county_avg <- WA_demographics %>%
  group_by(county_name) %>%
  summarize(
    num_tracts = n(),
    avg_pct_white = round(mean(pct_white, na.rm = TRUE), 2),
    avg_pct_black = round(mean(pct_black, na.rm = TRUE), 2),
    avg_pct_hispanic = round(mean(pct_hispanic, na.rm = TRUE), 2)
  )

kable(county_avg,
      col.names = c("County", "Count of Tracts", "Avg % White", "Avg % Black", "Avg % Hispanic"),
      caption = "Average Demographics by County",
      digits = 2,
      format.args = list(big.mark = ","))

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group

# Create a nicely formatted table of your results using kable()
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see if some communities have less reliable data.

**Requirements:**
- Calculate MOE percentages for each demographic variable
- Flag tracts where any demographic variable has MOE > 15%
- Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
MOE_WA_demographics <- WA_demographics %>%
 mutate(
   MOE_pct_White = (whiteM/whiteE) *100, 
   MOE_pct_Black = (blackM/blackE)*100, 
   MOE_pct_Hispanic= (hispanicM/hispanicE)*100

 )
# Create a flag for tracts with high MOE on any demographic variable
# Use logical operators (| for OR) in an ifelse() statement

MOE_WA_demographics <- MOE_WA_demographics %>%
  mutate(MOE_flag = ifelse(MOE_pct_White > 15 | MOE_pct_Black > 15 | MOE_pct_Hispanic > 15, "Yes", "No"))

# Create summary statistics showing how many tracts have data quality issues
MOE_WA_demographics %>%
  group_by(MOE_flag) %>%
  summarize(
    num_tracts = n()
  )
  
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
# Calculate average characteristics for each group:
# - population size, demographic percentages

pattern_analysis <- MOE_WA_demographics %>%
  group_by(MOE_flag) %>%
  summarize(
    num_tracts = n(),
    avg_population = round(mean(total_popE, na.rm = TRUE), 0),
    avg_pct_white = round(mean(pct_white, na.rm = TRUE), 2),
    avg_pct_black = round(mean(pct_black, na.rm = TRUE), 2),
    avg_pct_hispanic = round(mean(pct_hispanic, na.rm = TRUE), 2)
  )

kable(pattern_analysis,
      col.names = c("High MOE Flag", "Number of Tracts", "Avg Population", "Avg % White", "Avg % Black", "Avg % Hispanic"),
      caption = "Data Patterns by Tract Characteristics",
      digits = 2,
      format.args = list(big.mark = ","))

# Use group_by() and summarize() to create this comparison
# Create a professional table showing the patterns


```

**Pattern Analysis:** This table shows an interesting pattern, if we can call it a pattern at all. The pattern analysis shows that all 689 census tracts in Washington State were flagged for high MOE on at least one variable (White, Black, or Hispanic), indicating that data quality issues are widepread across the state instead of being isolated to particular areas. The table shows very small shares of non-white populations; when a  tract has very few Black or Hispanic residents, even a small absolute MOE could be reflected as a large percentage, making the 15% threshold we set very difficult or even impossible to meet across Washington's predominantly white communities.

When we consider equity issues that could arise from algorithmic bias, it is very important to consider if the thresholds we apply are considerate of the demopgraphic make-up in the tracts. What counts as unreliable data for a majority minority urban tract will end up being very different from what it means for a rural predominantly white one. In future analyses, it would be more equitable to explore MOE thresholds that are specific to population size, and demographic make-up. 


# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from all four analyses.

**Executive Summary Requirements:**
1. **Overall Pattern Identification**: What are the systematic patterns across all your analyses?

Across all four analyses, there is a consistent pattern where data reliability in Washington State is quite uneven and seemingly tied to population size. County-level income data shopwed that that the least reliable estimates (MOE % > 10%) are concentrated in smaller rural counties like Garfield, Wahkiakum, and Pend Oreille. Now diving deeper, at the tract level, this problem is exacerbated further with all 689 tracts flagged for high uncertainty on at least one variable (White, Black, or Hispanic). This show the fragmented reality of estimates with extremely different levels of confidence.

2. **Equity Assessment**: Which communities face the greatest risk of algorithmic bias based on your findings?

The communitiies facing the greatest risk of algorithmic bias tend to be the small rural counties with low and moderate incomes, such as Garfield and Ferry Counties, where data uncertainty is highest. These communities would be most harmed by algorithms that treats uncertain estimates with the same confidence threshold as reliable ones, resulted in misrepresented need for social services and improper distribution of funding as a result. Additionally, minority communities face increasing risk because data is the least statistically confident from small population counts. Therefore, algorithms that use racial breakdowns will probably be working with unreliable information for the communities that would most need equitable representation in policy decisions.

3. **Root Cause Analysis**: What underlying factors drive both data quality issues and bias risk?

The underlying factor of these data quality issues is merely the nature of the ACS, where, unlike a full census count, the ACS takes a sample the population, which means already-small communities will produce even fewer responses and thus less statistically stable estimates. Evidently, this disadvantages rural and more sparse populations. Further, considering Washington is predominantly white means that Black and Hispanic residents make up very small shares of most tracts, causing MOE percentages to increase substantially. 

4. **Strategic Recommendations**: What should the Department implement to address these systematic issues?

The Department of Human Services should not have a uniform algorithmic system without first looking at and analyzing data reliability. To illustrate, high confidence counties like King County might be more appropriate for algorithmic decision-making, but low confidence counties like Garfield should require mandatory human review. If possible and/or feasible, the Department should also look into collecting supplemental data in rural and minority communities, which tends to be regions with weaker estimates. Any MOE thresholds used in future analyses should be based on the demographic and geographic context of each community rather than applied uniformly statewide.


**Executive Summary:**


Across all four analyses, data reliability in Washington State has shown to be very uneven. Even more so, it has consistently been tied to population size, with the least reliable estimates in small  counties and even higher unreliability at the tract level  (all 689 tracts were flagged for high uncertainty). The communities facing the greatest risk of algorithmic bias are small, lower-income rural counties (e.x., Garfield and Ferry), AND minority communities. Clearly, this means any algorithm using racial breakdowns would be using weakest data for communities most in need of equitable representation. Because the ACS is a sample survey rather than a full count, it structurally disadvantages already-small populations and causes MOE percentages for minority groups to increase substantially in a predominantly white state, such as Washington. In response, the Department of Human Services should move away from a uniform algorithmic approach and instead build data reliability directly into decision-making, which includes human-based review initially for low confidence counties. In addition, it would be beneficial to puruse additional data collection of underrepresented communities, while using MOE thresholds specific to demographic and geographical contexts of each community rather than applying a standard statewide.


## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

final_recommendations <- WA_reliability %>%
  select(county_name, med_hh_incE, MOE_pct, reliability) %>%
  mutate(algorithm_recommendation = case_when(
    reliability == "High Confidence" ~ "Safe for algorithmic decisions",
    reliability == "Moderate Confidence" ~ "Use with caution - monitor outcomes",
    reliability == "Low Confidence" ~ "Requires manual review or additional data"
  ))

# Format as a professional table with kable()

kable(final_recommendations,
      col.names = c("County", "Median Income", "MOE %", "Reliability", "Algorithm Recommendation"),
      caption = "County-Level Algorithm Implementation Framework",
      digits = 2,
      format.args = list(big.mark = ","))

#Specific breakdown

# High Confidence Counties
high_conf <- final_recommendations %>%
  filter(reliability == "High Confidence") %>%
  kable(col.names = c("County", "Median Income", "MOE %", "Reliability", "Recommendation"),
        caption = "Counties Suitable for Algorithmic Implementation",
        digits = 2,
        format.args = list(big.mark = ","))

# Moderate Confidence Counties
mod_conf <- final_recommendations %>%
  filter(reliability == "Moderate Confidence") %>%
  kable(col.names = c("County", "Median Income", "MOE %", "Reliability", "Recommendation"),
        caption = "Counties Requiring Additional Oversight",
        digits = 2,
        format.args = list(big.mark = ","))

# Low Confidence Counties
low_conf <- final_recommendations %>%
  filter(reliability == "Low Confidence") %>%
  kable(col.names = c("County", "Median Income", "MOE %", "Reliability", "Recommendation"),
        caption = "Counties Needing Alternative Approaches",
        digits = 2,
        format.args = list(big.mark = ","))

high_conf
mod_conf
low_conf
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to the department.

1. **Counties suitable for immediate algorithmic implementation:** A significant share of the the counties were considered high confidence. King County and other high confidence counties are better fit for algorithmic decision-making because their income estimates carry MOE percentages less than 5%, meaning the data is stable enough to support such automated decisions without any major risk of misallocation. 

2. **Counties requiring additional oversight:** Moderate confidence counties like Asotin, Grant, and Jefferson (total of 15 counties), should use algorithms with caution. This could mean having human reviewers periodically auditing outputs to find instances where estimates may be leading to inconsistent or inequitable funding decisions.

3. **Counties needing alternative approaches:** Becasue the threshold used earlier was not updated, there are still only 3 counties (Garfield, Wahkiakum, and Pend Oreille) that would require any sort of human-interaction rather than an algorithm on its own. Instead, the Department should consider, as mentioned earlier,  community outreach for supplemental data, or possibly other data (if feasible) from local agencies to build more reliability.

## Questions for Further Investigation

1. Are there clusters of low confidence tracts within counties, and if so, do they correspond to known Indigenous reservations, migrant worker communities, and/or other communities that have been historically under-counted?

2. How do these income data reliability issues play a role in other variables like housing instability or unemployment? Would incorporating varied data sources minimize the risk of algorithmic bias in funding decisions?

# Technical Notes

**Data Sources:** 
- U.S. Census Bureau, American Community Survey 2018-2022 5-Year Estimates
- Retrieved via tidycensus R package on [date]

**Reproducibility:** 
- All analysis conducted in R version [your version]
- Census API key required for replication
- Complete code and documentation available at: [your portfolio URL]

**Methodology Notes:**
Briefly, counties were selected to represent the full range of data reliability, where King County represents high confidence, Asotin represents moderate confidence, and Garfield represents low confidence. MOE percentages were calculated as the margin of error divided by the estimate, multiplied by 100.

**Limitations:**
There were a few notable limitaitons of this analysis. First, county-level data covers all of Washington State, but the tract-level demographic analysis was based on only three counties- the demographic patterns may not be generalized to the full state. Second, the uniform 15% MOE threshold applied across all demographic groups could not account for structural differences in population size, which as demonstrated, resulted in standardized flagging across all tracts and, thus, limits the actual usefulness of the flag (ex.yes/no). Further, this analysis relies only on ACS 5-year estimates from 2022 and cannot speak to whether these patterns are changing over time. The median household income is used as the primary measure of community need, but income alone is not the only proxy for poverty and resource scarcity since it does not consider things like wealth, debt,or access to public services. Finally, communities that are most under-counted in the census, such as undocumented residents, un-housed peoples, and people living in housing insecurity, are underrepresented here, in which the analysis happens to systematically underestimate need in the most vulnerable communities. 

---

## Submission Checklist

Before submitting your portfolio link on Canvas:

- [ ] All code chunks run without errors
- [ ] All "[Fill this in]" prompts have been completed
- [ ] Tables are properly formatted and readable
- [ ] Executive summary addresses all four required components
- [ ] Portfolio navigation includes this assignment
- [ ] Census API key is properly set 
- [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself. Your assignment should be accessible at `your-portfolio-url/labs/lab_1/your_file_name.html`