---
title: "Philadelphia Housing Price Prediction"
subtitle: "Improving Property Tax Assessments"
author: "Jams & Co | Jenny, Arzoo, Maude, Shubhanga "
date: "17 March 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    code-line-numbers: true
    incremental: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#|echo: false
#|eval: false

library(car)
library(caret)
library(ggplot2)
library(kableExtra)
library(knitr)
library(lubridate) #used to clean housing data
library(modelsummary)
library(patchwork)
library(RColorBrewer)
library(scales)
library(sf)
library(stargazer) #for summaries
library(tidycensus)
library(tidyverse)
library(tigris)
```

```{r}
#|echo: false
#|eval: false
#|message: false
#|warning: false
#|results: 'hide'

sale_price_byblockgroup_geo <- st_read("data/data_clean/sale_pricebyblockgroup_geo.geojson", quiet = TRUE)
housing_census_geo_no_na <- read.csv("data/data_clean/housing_census_geo_no_na.csv")
redlining_phl <- st_read("data/data_clean/redlining_phl.shp", quiet = T)
dat_load <- read_csv("data/data_clean/dat.csv")
n_hoods <- st_read("data/data_clean/n_hoods.shp", quiet = T)

```

# The Challenge

Home Sales Price Prediction for Sales Tax: Philadelphia Office of
Property Assessment

## Data Sources

-   **City of Philadelphia - for Housing Data, n= 15,307, 2023 - 2024**

    Sales Price, Total Livable Area, No. of Bathrooms, Age

-   **American Community Survey (2024) - for demographic data**

    Education, Poverty, Race, Vacancy

-   **Open Data Philly - for spatial features**

    Crime, Transit Stops, Parks

## Home Sale Price by Block Group

Northwest and Center City areas have the highest home sale prices

```{r}
#|echo: false
#|eval: false
#Avg sales price per block group
ggplot() +
  geom_sf(data = sale_price_byblockgroup_geo, aes(fill = mean_price)) +
  scale_fill_viridis_c(
    name = "Mean Sale Price ($)",
    option = "inferno",
    na.value = "grey",
    labels = scales::dollar
  ) +
  labs(
    title = "Average Residential Sale Price by Block Group",
    subtitle = "Philadelphia, 2023–2024 Sales",
    caption = "Source: OPA Sales Data"
  ) +
  theme_void()
```

## What Drives Prices?

Historic redlining grades overlap with current home sale prices

```{r}
#|echo: false
#|eval: false
#both maps combined with lower transparncy for avg sales/block group
ggplot() +
  geom_sf(data = sale_price_byblockgroup_geo, aes(fill = mean_price), 
          color = "white", size = 0.01, alpha = 0.45) +
  geom_sf(data = redlining_phl, aes(color = as.factor(grade)), 
          fill = NA, size = 0.8) +
  scale_fill_viridis_c(
    name = "Mean Sale Price ($)",
    option = "inferno",
    na.value = "grey",
    labels = scales::dollar
  ) +
  scale_color_manual(
    name = "HOLC Grade",
    values = c("0" = "grey", "1" = "green", "2" = "blue", "3" = "yellow", "4" = "red"),
    labels = c("1" = "A", "2" = "B", "3" = "C", "4" = "D")
  ) +
  labs(
    title = "Average Residential Sale Price by Block Group with Redlining Grades",
    subtitle = "Philadelphia, 2023–2024 Sales",
    caption = "Source: OPA Sales Data; HOLC Redlining Boundaries"
  ) +
  theme_void()
```

## What Drives Prices?

```{r}
#|echo: false
#|eval: false
#|message: false
#|warning: false
#|results: 'hide'



#scatterplots for correlation between sale price and structural data
housing_census_geo_no_na <- read.csv("data/data_clean/housing_census_geo_no_na.csv")

#log sale price and number of bedrooms
p1 <- ggplot(data = housing_census_geo_no_na) +
  aes(x = number_of_bedrooms, y = log_sale_price) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = " Number of Bedrooms vs Log of Sales Price",
       x = "Number of Bedrooms",
       y = "Log Sales Price") +
  theme_minimal()

#log sale price and number of bathrooms 
p2 <- ggplot(data = housing_census_geo_no_na) +
  aes(x = number_of_bathrooms, y = log_sale_price) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Number of Bathrooms vs Log of Sales Price",
       x = "Number of Bathrooms",
       y = "Log Sales Price") +
  theme_minimal()

#log sale price and total liveable area
p3 <- ggplot(data = housing_census_geo_no_na) +
  aes(x = total_livable_area, y = log_sale_price) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = FALSE)+
  labs (title = "Total Livable Area vs Log of Sales Price",
        x = "Total Livable Area",
         y = "Log of Sales Price") +
theme_minimal()

housing_census_geo_no_na$year_built <- as.numeric(housing_census_geo_no_na$year_built)


#log sale price and year built

#log sale price and share of residents with at least a bachelor's degree
p5 <-  ggplot(data = housing_census_geo_no_na) +
  aes(x = BACH_PROP, y = log_sale_price) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = " Bachelor's Degree vs Log of Sales Price",
       x = "Bachelor's Degree Proportion",
       y = "Log of Sales Price") +
  theme_minimal()

 p4 <- ggplot(data = dat_load) +
  aes(x = age, y = log_sale_price.x) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = " Age of building vs Log of Sales Price",
       x = "Age in Years",
       y = "Log of Sales Price") +
  theme_minimal()

#google and claude (generate the code) and r blogs assisted for beautification and fitting
combined_struc <- (p1 + p2 + p3 + p4 + p5)+
  plot_layout(ncol = 2) &
  theme(plot.title = element_text(size = 8),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 6))
combined_struc

```

## Model Comparison

```{r}
#|echo: false
#|eval: false
#|message: false
#|warning: false
#|results: 'hide'

ctrl <- trainControl(method = "cv", number = 10)


#Model 1: Structural only
structural <- lm(log_sale_price.x ~ age+ I(age^2) + number_of_bathrooms.x + 
            log_liv_area, 
          data = dat_load)


#Model 2: + Demographic
demographic <- lm(log_sale_price.x ~  age+ I(age^2) + number_of_bathrooms.x + 
            log_liv_area + BACH_PROP + WHITE_PROP + Pov_Prop + 
            Vacancy_Rate, 
          data = dat_load)


#Model 3: + Spatial
spatial <- lm(log_sale_price.x ~  age+ I(age^2) + number_of_bathrooms.x + 
            log_liv_area + BACH_PROP + WHITE_PROP + Pov_Prop + 
            Vacancy_Rate + log_crime_score + near_transit + parks_400, 
          data = dat_load)

#Model 4: + Fixed Effects (chosen model)
FE <- lm(log_sale_price.x ~ as.factor(name_cv) +  age+ I(age^2) + 
            number_of_bathrooms.x + log_liv_area + BACH_PROP + WHITE_PROP + 
            Pov_Prop + Vacancy_Rate + log_crime_score + near_transit + 
            parks_400, 
          data = dat_load)



#setting up 10-fold cross validation - splitting data into 10 equal chunks
ctrl <- trainControl(method = "cv", number = 10)

# CV for all 4 models
cv_str <- train(log_sale_price.x ~ age + I(age^2) + number_of_bathrooms.x + 
                 log_liv_area,
               data = dat_load, method = "lm", trControl = ctrl)

cv_demo <- train(log_sale_price.x ~  age + I(age^2)  + number_of_bathrooms.x + 
                 log_liv_area + BACH_PROP + WHITE_PROP + Pov_Prop + 
                 Vacancy_Rate,
               data = dat_load, method = "lm", trControl = ctrl)

cv_spat <- train(log_sale_price.x ~  age + I(age^2)  + number_of_bathrooms.x + 
                 log_liv_area + BACH_PROP + WHITE_PROP + Pov_Prop + 
                 Vacancy_Rate + log_crime_score + near_transit + parks_400,
               data = dat_load, method = "lm", trControl = ctrl)

cv_fe<- train(log_sale_price.x ~ as.factor(name_cv) +  age + I(age^2)  + 
                 number_of_bathrooms.x + log_liv_area + BACH_PROP + WHITE_PROP + 
                 Pov_Prop + Vacancy_Rate + log_crime_score + near_transit + 
                 parks_400,
               data = dat_load, method = "lm", trControl = ctrl)













#Results table
cv_results <- tibble(
  Model   = c("Structural", "+ Demographic", "+ Spatial", "+ Fixed Effects"),
  RMSE    = c(cv_str$results$RMSE, cv_demo$results$RMSE, 
              cv_spat$results$RMSE, cv_fe$results$RMSE),
  MAE     = c(cv_str$results$MAE,  cv_demo$results$MAE,  
              cv_spat$results$MAE,  cv_fe$results$MAE),
  R2      = c(cv_str$results$Rsquared, cv_demo$results$Rsquared, 
              cv_spat$results$Rsquared, cv_fe$results$Rsquared)
) %>%
  kable(digit = 3) %>%
  kable_styling(full_width = F, position = "center")

cv_results

```

## Top Predictors

1.  **Livable area**: A 10% larger home sells for \~6.3% more
2.  **Number of bathrooms**: Each additional bathroom adds \~19% to sale
    price
3.  **Race:** A block group 10 percentage points whiter is associated
    with homes selling for 4.1% more
4.  **Parks within 400m**: Neighborhoods with double the park access see
    prices \~2.8% higher
5.  **Log crime score**: A 10% higher crime score is associated with
    0.47% lower sale prices

## Model Performance

**Final Model with Neighborhood Fixed Effects**

1.  RMSE to 0.3905 and explaining variance up to 70%.
2.  Median Absolute Error is USD 42,994 against a median sale price of
    USD 250,000.

## Model Performance

```{r}
#calculating residuals
dat <- dat_load %>%
  mutate(residual = log_sale_price.x - predicted)

#aggregating to neighborhood level
residuals_by_hood <- dat_load %>%
  group_by(name_cv) %>%
  summarise(mean_residual = mean(residual, na.rm = TRUE),
            n = n()) %>%
  ungroup()

#joining back to neighborhood shapefile
n_hoods_residuals <- n_hoods %>%
  left_join(residuals_by_hood, by = c("Name" = "name_cv"))

#mapping it
ggplot() +
  geom_sf(data = n_hoods_residuals, 
          aes(fill = mean_residual)) +
  scale_fill_gradient2(
    low      = "lightgreen",
    mid      = "orange", 
    high     = "tomato",
    midpoint = 0,
    name     = "Mean Residual"
  ) +
  labs(
    title    = "Where Does Our Model Struggle?",
    subtitle = "Mean Prediction Residual by Neighborhood",
    caption  = "Red = model underestimates price | Green = model overestimates price"
  ) +
  theme_void()

```

## Limitations & Next Steps

-   Limitation: One neighborhood where prices are systematically
    underestimated is Center City East, while the neighborhood where
    prices are overestimated is Fairhill, which needs to be further
    investigated.

-   Next Steps: Jams & Co intends to enhance the model by conducting
    other model diagnostics, by analyzing factors such as spatial
    autocorrelation.

## Recommendations

-   Our model is a holistic and comprehensive predicive model for home
    sale prices, and we recommend it be used to support property tax
    determination

-   We recommend supplementary screening and surveying for neighborhoods
    that are systematically under and over predicted by the model.

# Thank you.
